A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s

### 2018-11-07 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
add references

While the density of transistors on silicon continues to increase
following Moore's Law (albeit more slowly than in the past), power
density constraints have limited the clock speed of processors for more
than a decade. This has effectively stalled any progress in the
processing capacity of a single CPU core. Instead, increases in
potential processing capacity come from increases in the core count of
CPUs and wide CPU registers. Alternative processing architectures have
become more commonplace. These range from the many-core architecture
based on standard x86\_64 cores to numerous alternatives such as GPUs.
For GPUs, the processing model is very different, allowing a much
greater fraction of the die to be dedicated to arithmetic calculations,
but at a price in programming difficulty and memory handling for the
developer that tends to be specific to each processor generation.
Further developments may even see the use of FPGAs for more
general-purpose tasks. Fully exploiting these evolutions requires a
shift in programming model to one based on \emph{concurrency}.

Even with the throttling of clock speed to limit power consumption,
power remains a major issue. Low power architectures are in huge demand.
At one level this might challenge the dominance of x86\_64 by simply
replacing it with, for example, AArch64 devices that may achieve lower power
costs for the scale of HEP computing needs than Intel has achieved with its
Xeon architecture. More extreme is an architecture that would see
specialised processing units dedicated to particular tasks, but with
possibly large parts of the device switched off most of the time,
so-called dark silicon~\cite{IntelCSA}.

@@

Particle physics is no longer alone in facing these massive data
challenges. Experiments in other fields, from astronomy to genomics,
will produce huge amounts of data in the future, and will need to
overcome the same challenges that we face, i.e., massive data handling and
efficient scientific programming. Establishing links with these fields
has already started. Additionally, interest from the computing science
community in solving these data challenges exists, and mutually
beneficial relationships would be possible where there are genuine
research problems that are of academic interest to that community and
provide practical solutions to ours. The efficient processing of massive
data volumes is also a challenge faced by industry, in particular the
internet economy, which developed novel and major new technologies under
the banner of \emph{Big Data} that may be applicable to our use cases.

@@

\paragraph{Fast Simulation} The work on Fast Simulation is also
accelerating with the objective of producing a flexible framework that
permits Full and Fast simulation to be combined for different particles
in the same event. Various approaches to Fast Simulation are being tried
all with the same goal of saving computing time, under the assumption
that it is possible to improve time performance without an unacceptable
loss of physics accuracy. There has recently been a great deal of
interest in the use of Machine Learning in Fast Simulation, most of
which has focused on the use of multi-objective regression and
generative adversarial networks (GANs). Since use of GANs allows for
non-parametric learning in cases such as calorimetric shower
fluctuations, it is a promising avenue for generating non-Gaussian and
highly correlated physical effects. This is an obvious area for future
expansion and development, as it is currently in its infancy.

@@

Current experiments have moved towards smaller, but still
flexible, tiered data formats. These tiers are typically based on the
ROOT~\cite{Brun1996} file format and constructed to
facilitate both skimming of interesting events and the selection of
interesting pieces of events by individual analysis groups or through
centralised analysis processing systems. Initial implementations of
real-time analysis systems are in use within several experiments. These
approaches remove the detector data that typically makes up the raw data
tier kept for offline reconstruction, and keep only final analysis
objects~\cite{Aaij2016,ATL-DAQ-PUB-2017-003,Khachatryan:2149625}.

@@

Over the past 20 years the HEP community has developed and gravitated
around a single analysis ecosystem based on ROOT~\cite{Brun1996}. 
ROOT is a general-purpose object
oriented framework that addresses the selection, integration,
development, and support of a number of foundation and utility class
libraries that can be used as a basis for developing HEP application
codes. The added value to the HEP community is that it provides an
integrated and validated toolkit, and its use encompasses the full event
processing chain; it has a major impact on the way HEP analysis is
performed. This lowers the hurdle to start an analysis, enabling the
community to communicate using a common analysis language, as well as
making common improvements as additions to the toolkit quickly become
available. The ongoing ROOT programme of work addresses important new
requirements, in both functionality and performance, and this is given
a high priority by the HEP community.

@@

One new model of data analysis, developed outside of HEP, maintains the
concept of sequential reduction, but mixes interactivity with
batch processing. These exploit new cluster management systems, most
notably Apache Spark, which uses open-source tools contributed both by
industry and the data-science community. Other products implementing the
same analysis concepts and workflows are emerging, such as TensorFlow,
Dask, Pachyderm, Blaze, Parsl, and Thrill. This approach can complement
the present and widely adopted Grid processing of datasets. It may
potentially simplify the access to data and the expression of
parallelism, thereby improving the exploitation of cluster resources.

@@

By 2020:
\begin{itemize}
\item
  \emph{Particle identification and particle properties:} in calorimeters or
  time projection chambers (TPCs), where the data can be represented as
  a 2D or 3D image (or even in 4D, including timing information), the
  problems can be cast as a computer vision task. Deep Learning (DL),
  one class of ML algorithm, in which neural networks are used to
  reconstruct images from pixel intensities, is a good candidate to
  identify particles and extract many parameters. Promising DL
  architectures for these tasks include convolutional, recurrent, and
  adversarial neural networks. A particularly important application is
  to Liquid Argon TPCs (LArTPCs), which is the chosen detection
  technology for DUNE, the new flagship experiment in the neutrino programme. 
  A proof of
  concept and comparison of DL architectures should be finalised by
  2020. Particle identification can also be explored to tag the flavour
  of jets in collider experiments (e.g., so-called b-tagging). The
  investigation of these concepts, which connect to Natural Language
  Processing, has started at the LHC and is to be pursued on the same
  timescale.
\item
  \emph{ML middleware and data formats for offline usage:} HEP relies 
  on the ROOT format for its data, wheras the ML community
  has developed several other formats, often associated with specific ML
  tools. A desirable data format for ML applications should have the
  following attributes: high read-write speed for efficient training,
  sparse readability without loading the entire dataset into RAM,
  compressibility, and widespread adoption by the ML community. The thorough
  evaluation of the different data formats and their impact on ML
  performance in the HEP context must be continued, and it is necessary
  to define a strategy for bridging or migrating HEP formats to the
  chosen ML format(s), or vice-versa.
\item
  \emph{Computing resource optimisations:} managing large
  volume data transfers is one
  of the challenges facing current computing facilities. Networks 
  play a crucial role in data exchange and so a
  network-aware application layer may significantly improve experiment
  operations. ML is a promising technology to identify anomalies in
  network traffic, to predict and prevent network congestion, to detect
  bugs via analysis of self-learning networks, and for WAN path
  optimisation based on user access patterns.
\item
  \emph{ML as a Service (MLaaS):} current cloud providers rely on a MLaaS model
  exploiting interactive
  machine learning tools in order to make efficient use of resources, however,
  this is not yet widely used in HEP. HEP
  services for interactive analysis, such as CERN's Service for
  Web-based Analysis, SWAN~\cite{SWAN}, may play an important role in adoption
  of machine learning tools in HEP workflows. In order to use these tools
  more efficiently, sufficient and appropriately tailored hardware and
  instances other than SWAN will be identified.
\end{itemize}

By 2022:
\begin{itemize}
\item
  \emph{Detector anomaly detection:} data taking is
  continuously monitored by physicists taking shifts to monitor and
  assess the quality of the incoming data, largely using reference
  histograms produced by experts. A whole class of ML algorithms called anomaly detection
  can be useful for automating this important task. Such unsupervised algorithms are able
  to learn from data and produce an alert when deviations are observed.
  By monitoring many variables at the same time, such algorithms are
  sensitive to subtle signs forewarning of imminent failure, so that
  pre-emptive maintenance can be scheduled. These techniques are already
  used in industry.
\item
  \emph{Simulation:} recent progress in high fidelity fast generative models,
  such as Generative Adversarial Networks (GANs) and Variational
  Autoencoders (VAEs), which are able to sample high dimensional feature
  distributions by learning from existing data samples, offer a
  promising alternative for Fast Simulation. A simplified first attempt at
  using such techniques in simulation saw orders of magnitude increase
  in speed over existing Fast Simulation techniques, but has not yet
  reached the required accuracy~\cite{CaloGAN}.
\item
  \emph{Triggering and real-time analysis:} one of the challenges is the
  trade-off in algorithm complexity and performance under strict
  inference time constraints. To deal with the increasing event
  complexity at HL-LHC, the use of sophisticated ML
  algorithms will be explored at all trigger levels, building on the pioneering work of
  the LHC collaborations. A critical part of this work will be to
  understand which ML techniques allow us to maximally exploit
  future computing architectures.
\item
  \emph{Sustainable Matrix Element Method (MEM):} MEM is a powerful
  technique that can be utilised for making measurements of physical model
  parameters and direct searches for new phenomena. As it is
  very computationally intensive its use in HEP is limited.
  Although the use of neural networks for numerical integration is not new, 
  it is a
  technical challenge to design a network sufficiently rich
  to encode the complexity of the ME calculation for a given process
  over the phase space relevant to the signal process. Deep Neural
  Networks (DNNs) are good candidates~\cite{Bendavid2017,BendavidIML2017}.
\item
  \emph{Tracking:} pattern recognition is always a computationally challenging
  step. It becomes a huge challenge in the HL-LHC environment.
  Adequate ML techniques may provide a solution that scales linearly
  with LHC intensity. Several efforts in the HEP community have started
  to investigate ML algorithms for track pattern recognition on
  many-core processors.
\end{itemize}

@@

As outlined in Section \ref{software-and-computing-challenges}, 
huge resource requirements are anticipated for
HL-LHC running. These need to be deployed and managed across the WLCG
infrastructure, which has evolved from the original ideas on deployment
before LHC data-taking started~\cite{MONARC}, to be a mature and
effective infrastructure that is now exploited by LHC experiments.
Currently, hardware costs are dominated by disk storage, closely
followed by CPU, followed by tape and networking. Naive estimates of
scaling to meet HL-LHC needs indicate that the current system would need
almost an order of magnitude more resources than will be available from
technology evolution alone. In addition, other initiatives such as Belle
II and DUNE in particle physics, but also other science projects such as
SKA, will require a comparable amount of resources on the same
infrastructure. Even anticipating substantial software improvements, the
major challenge in this area is to find the best configuration for
facilities and computing sites that make HL-LHC computing feasible. This
challenge is further complicated by substantial regional differences in
funding models, meaning that any solution must be sensitive to these
local considerations to be effective.

@@

In recent years, video game engines, such as Unity or the Unreal Engine,
have become particularly popular in the game and architectural
visualisation industry. They provide very sophisticated graphics engines
and offer a lot of tools for user interaction, such as menu systems or
native handling of VR devices. They are well supported by industry and
tend to have a long lifespan (Unreal Engine is now  20 years old and
is still very popular). However, such engines are meant to be used as
development frameworks and their usage in HEP code is not always
evident. Code should be developed within them, while in HEP
framework-based applications we often want to use graphics libraries
that can be integrated in existing code. A number of HEP collaborations
have started experimenting in building event display tools with such
engines, among them Belle II and ATLAS, but their use is currently
limited to the display of simplified data only.

@@

Although non-X509 federated identity provides a potential solution for our
challenges in Authentication, Authorisation should continue to be
tightly controlled by the HEP community. Enabling Virtual Organisation
(VO) membership for federated credentials and integrating such a
workflow with existing identity vetting processes is a major topic
currently being worked on, in particular within the WLCG community.
Commercial clouds and HPC centres have fundamentally different access
control models and technologies from our grid environment. We shall need
to enhance our access control model to ensure compatibility and
translate our grid-based identity attributes into those consumable by
such services.

@@

New pedagogical methods, such as active training and peer training, that
are complementary to schools or topical tutorials, also deserve more
attention. Online material can be shared by a student and a teacher to
provide the exchange of real examples and practical exercises. For
example, notebook technologies, such as Jupyter, support embedding of
runnable code and comments into the same document. The initial material
can be easily enriched by allowing other students and experts to add
comments and more examples in a collaborative way. The HSF started to
experiment with this approach with WikiToLearn~\cite{WikiToLearn}, a
platform developed in Italy outside HEP that promotes this kind of
training and collaborative enrichment of the training material. Projects
such as ROOT~\cite{Brun1996} have also started to provide some
training material based
on notebooks.

@@

A model that emerged in recent years as a very valuable means of sharing
expertise is to use Question and Answer (Q\&A) systems, such as Stack
Overflow. A few such systems are run by experiments for their own needs,
but this is not necessarily optimal, as the value of these services is
increased by a large number of contributors with diverse backgrounds.
Running a cross-experiment Q\&A system has been discussed, but it has
not yet been possible to converge on a viable approach, both technically
and because of the effort required to run and support such a service.

>>>>

While the density of transistors on silicon continues to increase
following Moore's Law (albeit more slowly than in the past), power
density constraints have limited the clock speed of processors for more
than a decade~\cite{NAP12980}. This has effectively stalled any progress in the
processing capacity of a single CPU core. Instead, increases in
potential processing capacity come from increases in the core count of
CPUs and wide CPU registers. Alternative processing architectures have
become more commonplace. These range from the many-core architecture
based on standard x86\_64 cores to numerous alternatives such as GPUs.
For GPUs, the processing model is very different~\cite{Cook:2012:CPD:2430671}, allowing a much
greater fraction of the die to be dedicated to arithmetic calculations,
but at a price in programming difficulty and memory handling for the
developer that tends to be specific to each processor generation.
Further developments may even see the use of FPGAs for more
general-purpose tasks. Fully exploiting these evolutions requires a
shift in programming model to one based on \emph{concurrency}.

Even with the throttling of clock speed to limit power consumption,
power remains a major issue. Low power architectures are in huge demand.
At one level this might challenge the dominance of x86\_64 by simply
replacing it with, for example, AArch64 devices that may achieve lower power
costs for the scale of HEP computing needs than Intel has achieved with its
Xeon architecture. More extreme is an architecture that would see
specialised processing units dedicated to particular tasks, but with
possibly large parts of the device switched off most of the time,
so-called dark silicon.

@@

Particle physics is no longer alone in facing these massive data
challenges. Experiments in other fields, from astronomy~\cite{10.1371/journal.pbio.1002195} 
to genomics~\cite{ijms18020412},
will produce huge amounts of data in the future, and will need to
overcome the same challenges that we face, i.e., massive data handling and
efficient scientific programming. Establishing links with these fields
has already started. Additionally, interest from the computing science
community in solving these data challenges exists, and mutually
beneficial relationships would be possible where there are genuine
research problems that are of academic interest to that community and
provide practical solutions to ours. The efficient processing of massive
data volumes is also a challenge faced by industry, in particular the
internet economy, which developed novel and major new technologies under
the banner of \emph{Big Data} that may be applicable to our use cases.

@@

\paragraph{Fast Simulation} The work on Fast Simulation is also
accelerating with the objective of producing a flexible framework that
permits Full and Fast simulation to be combined for different particles
in the same event. Various approaches to Fast Simulation are being tried
all with the same goal of saving computing time, under the assumption
that it is possible to improve time performance without an unacceptable
loss of physics accuracy. There has recently been a great deal of
interest in the use of Machine Learning in Fast Simulation, most of
which has focused on the use of multi-objective regression and
generative adversarial networks (GANs)~\cite{}. Since use of GANs allows for
non-parametric learning in cases such as calorimetric shower
fluctuations, it is a promising avenue for generating non-Gaussian and
highly correlated physical effects. This is an obvious area for future
expansion and development, as it is currently in its infancy.

@@

Current experiments have moved towards smaller, but still
flexible, tiered data formats. These tiers are typically based on the
ROOT file format and constructed to
facilitate both skimming of interesting events and the selection of
interesting pieces of events by individual analysis groups or through
centralised analysis processing systems. Initial implementations of
real-time analysis systems are in use within several experiments. These
approaches remove the detector data that typically makes up the raw data
tier kept for offline reconstruction, and keep only final analysis
objects~\cite{Aaij2016,ATL-DAQ-PUB-2017-003,Khachatryan:2149625}.

@@

Over the past 20 years the HEP community has developed and gravitated
around a single analysis ecosystem based on ROOT. 
ROOT is a general-purpose object
oriented framework that addresses the selection, integration,
development, and support of a number of foundation and utility class
libraries that can be used as a basis for developing HEP application
codes. The added value to the HEP community is that it provides an
integrated and validated toolkit, and its use encompasses the full event
processing chain; it has a major impact on the way HEP analysis is
performed. This lowers the hurdle to start an analysis, enabling the
community to communicate using a common analysis language, as well as
making common improvements as additions to the toolkit quickly become
available. The ongoing ROOT programme of work addresses important new
requirements, in both functionality and performance, and this is given
a high priority by the HEP community.

@@

One new model of data analysis, developed outside of HEP, maintains the
concept of sequential reduction, but mixes interactivity with
batch processing. These exploit new cluster management systems, most
notably Apache Spark~\cite{Armbrust:2015:SSR:2723372.2742797}, which uses open-source tools contributed both by
industry and the data-science community. Other products implementing the
same analysis concepts and workflows are emerging, such as TensorFlow~\cite{tensorflow2015-whitepaper},
Dask~\cite{dask_ref}, Pachyderm~\cite{thepachydermteam}, 
Blaze~\cite{blaze}, Parsl~\cite{babuji_yadu_2017_853492}, 
and Thrill~\cite{bingmann2016thrill}. This approach can complement
the present and widely adopted Grid processing of datasets. It may
potentially simplify the access to data and the expression of
parallelism, thereby improving the exploitation of cluster resources.

@@

By 2020:
\begin{itemize}
\item
  \emph{Particle identification and particle properties:} in calorimeters or
  time projection chambers (TPCs), where the data can be represented as
  a 2D or 3D image (or even in 4D, including timing information), the
  problems can be cast as a computer vision task. Deep Learning (DL),
  one class of ML algorithm, in which neural networks are used to
  reconstruct images from pixel intensities, is a good candidate to
  identify particles and extract many parameters. Promising DL
  architectures for these tasks include convolutional, recurrent, and
  adversarial neural networks. A particularly important application is
  to Liquid Argon TPCs (LArTPCs), which is the chosen detection
  technology for DUNE, the new flagship experiment in the neutrino programme. 
  A proof of
  concept and comparison of DL architectures should be finalised by
  2020. Particle identification can also be explored to tag the flavour
  of jets in collider experiments (e.g., so-called b-tagging). The
  investigation of these concepts, which connect to Natural Language
  Processing~\cite{Collobert:2011:NLP:1953048.2078186}, has started at the LHC and is to be pursued on the same
  timescale.
\item
  \emph{ML middleware and data formats for offline usage:} HEP relies 
  on the ROOT format for its data, wheras the ML community
  has developed several other formats, often associated with specific ML
  tools. A desirable data format for ML applications should have the
  following attributes: high read-write speed for efficient training,
  sparse readability without loading the entire dataset into RAM,
  compressibility, and widespread adoption by the ML community. The thorough
  evaluation of the different data formats and their impact on ML
  performance in the HEP context must be continued, and it is necessary
  to define a strategy for bridging or migrating HEP formats to the
  chosen ML format(s), or vice-versa.
\item
  \emph{Computing resource optimisations:} managing large
  volume data transfers is one
  of the challenges facing current computing facilities. Networks 
  play a crucial role in data exchange and so a
  network-aware application layer may significantly improve experiment
  operations. ML is a promising technology to identify anomalies in
  network traffic, to predict and prevent network congestion, to detect
  bugs via analysis of self-learning networks, and for WAN path
  optimisation based on user access patterns.
\item
  \emph{ML as a Service (MLaaS):} current cloud providers rely on a MLaaS model
  exploiting interactive
  machine learning tools in order to make efficient use of resources, however,
  this is not yet widely used in HEP. HEP
  services for interactive analysis, such as CERN's Service for
  Web-based Analysis, SWAN~\cite{SWAN}, may play an important role in adoption
  of machine learning tools in HEP workflows. In order to use these tools
  more efficiently, sufficient and appropriately tailored hardware and
  instances other than SWAN will be identified.
\end{itemize}

By 2022:
\begin{itemize}
\item
  \emph{Detector anomaly detection:} data taking is
  continuously monitored by physicists taking shifts to monitor and
  assess the quality of the incoming data, largely using reference
  histograms produced by experts. A whole class of ML algorithms called anomaly detection
  can be useful for automating this important task. Such unsupervised algorithms are able
  to learn from data and produce an alert when deviations are observed.
  By monitoring many variables at the same time, such algorithms are
  sensitive to subtle signs forewarning of imminent failure, so that
  pre-emptive maintenance can be scheduled. These techniques are already
  used in industry.
\item
  \emph{Simulation:} recent progress in high fidelity fast generative models,
  such as Generative Adversarial Networks (GANs)~\cite{NIPS2014_5423} and Variational
  Autoencoders (VAEs)~\cite{2013arXiv1312.6114K}, 
  which are able to sample high dimensional feature
  distributions by learning from existing data samples, offer a
  promising alternative for Fast Simulation. A simplified first attempt at
  using such techniques in simulation saw orders of magnitude increase
  in speed over existing Fast Simulation techniques, but has not yet
  reached the required accuracy~\cite{CaloGAN}.
\item
  \emph{Triggering and real-time analysis:} one of the challenges is the
  trade-off in algorithm complexity and performance under strict
  inference time constraints. To deal with the increasing event
  complexity at HL-LHC, the use of sophisticated ML
  algorithms will be explored at all trigger levels, building on the pioneering work of
  the LHC collaborations. A critical part of this work will be to
  understand which ML techniques allow us to maximally exploit
  future computing architectures.
\item
  \emph{Sustainable Matrix Element Method (MEM):} MEM is a powerful
  technique that can be utilised for making measurements of physical model
  parameters and direct searches for new phenomena. As it is
  very computationally intensive its use in HEP is limited.
  Although the use of neural networks for numerical integration is not new, 
  it is a
  technical challenge to design a network sufficiently rich
  to encode the complexity of the ME calculation for a given process
  over the phase space relevant to the signal process. Deep Neural
  Networks (DNNs) are good candidates~\cite{Bendavid2017,BendavidIML2017}.
\item
  \emph{Tracking:} pattern recognition is always a computationally challenging
  step. It becomes a huge challenge in the HL-LHC environment.
  Adequate ML techniques may provide a solution that scales linearly
  with LHC intensity. Several efforts in the HEP community have started
  to investigate ML algorithms for track pattern recognition on
  many-core processors.
\end{itemize}

@@

As outlined in Section \ref{software-and-computing-challenges}, 
huge resource requirements are anticipated for
HL-LHC running. These need to be deployed and managed across the WLCG
infrastructure, which has evolved from the original ideas on deployment
before LHC data-taking started~\cite{Aderholz:510694}, to be a mature and
effective infrastructure that is now exploited by LHC experiments.
Currently, hardware costs are dominated by disk storage, closely
followed by CPU, followed by tape and networking. Naive estimates of
scaling to meet HL-LHC needs indicate that the current system would need
almost an order of magnitude more resources than will be available from
technology evolution alone. In addition, other initiatives such as Belle
II and DUNE in particle physics, but also other science projects such as
SKA, will require a comparable amount of resources on the same
infrastructure. Even anticipating substantial software improvements, the
major challenge in this area is to find the best configuration for
facilities and computing sites that make HL-LHC computing feasible. This
challenge is further complicated by substantial regional differences in
funding models, meaning that any solution must be sensitive to these
local considerations to be effective.

@@

In recent years, video game engines, such as Unity~\cite{unity} or 
the Unreal Engine~\cite{Sanders:2016:IUE:3099885},
have become particularly popular in the game and architectural
visualisation industry. They provide very sophisticated graphics engines
and offer a lot of tools for user interaction, such as menu systems or
native handling of VR devices. They are well supported by industry and
tend to have a long lifespan (Unreal Engine is now  20 years old and
is still very popular). However, such engines are meant to be used as
development frameworks and their usage in HEP code is not always
evident. Code should be developed within them, while in HEP
framework-based applications we often want to use graphics libraries
that can be integrated in existing code. A number of HEP collaborations
have started experimenting in building event display tools with such
engines, among them Belle II and ATLAS, but their use is currently
limited to the display of simplified data only.

@@

Although federated identity provides a potential solution for our
challenges in Authentication, Authorisation should continue to be
tightly controlled by the HEP community. Enabling Virtual Organisation
(VO) membership for federated credentials and integrating such a
workflow with existing identity vetting processes is a major topic
currently being worked on, in particular within the WLCG community.
Commercial clouds and HPC centres have fundamentally different access
control models and technologies from our grid environment. We shall need
to enhance our access control model to ensure compatibility and
translate our grid-based identity attributes into those consumable by
such services.

@@

New pedagogical methods, such as active training and peer training, that
are complementary to schools or topical tutorials, also deserve more
attention. Online material can be shared by a student and a teacher to
provide the exchange of real examples and practical exercises. For
example, notebook technologies, such as Jupyter, support embedding of
runnable code and comments into the same document. The initial material
can be easily enriched by allowing other students and experts to add
comments and more examples in a collaborative way. The HSF started to
experiment with this approach with WikiToLearn~\cite{WikiToLearn}, a
platform developed in Italy outside HEP that promotes this kind of
training and collaborative enrichment of the training material. Projects
such as ROOT have also started to provide some
training material based
on notebooks.

@@

A model that emerged in recent years as a very valuable means of sharing
expertise is to use Question and Answer (Q\&A) systems, such as Stack
Overflow~\cite{Stackoverflow}. A few such systems are run by experiments for their own needs,
but this is not necessarily optimal, as the value of these services is
increased by a large number of contributors with diverse backgrounds.
Running a cross-experiment Q\&A system has been discussed, but it has
not yet been possible to converge on a viable approach, both technically
and because of the effort required to run and support such a service.
